---
title: "AutohierarchyM: automatic hierarchy construction using marker genes I"
output: html_notebook
---

```{r, message=FALSE, warning=FALSE, results="hide"}
#devtools::install_github("khodosevichlab/CellAnnotatoR")
#devtools::install_github("hms-dbmi/sccore", ref="dev", force=TRUE)
#devtools::install_github('YuLab-SMU/ggtree') #package ‘aplot’ is not available (for R version 3.5.1)
#install.packages("pROC") #For AUC
#devtools::load_all("~/CellAnnotatoR") # Every time one changes the functions in R folder, one needs to reload the package.
```

```{r, message=FALSE, warning=FALSE, results="hide"}
library(sccore)
library(ggplot2)
library(magrittr)
library(Matrix)
library(pbapply)
library(dplyr)
library(pagoda2)
library(readr)
library(CellAnnotatoR)

theme_set(theme_bw())
```


# Automatic hierarchy construction using marker genes I

## Preparation

#### Build an automatic hierarchy from the clusters at the lowest level of the manual hierarchy ann_by_level$annotation$l3 using deriveHierararchy.
```{r}
p2 <- read_rds("~/data/separability/mouse_visp_marker_selection/p2.rds")
ann.by.level <- readRDS("~/data/separability/mouse_visp_marker_selection/ann_by_level.rds") #names(ann.by.level): "annotation" #annotation using the previous tree and markers    "scores" #S(t)   "annotation.filt"   
hierarchy <- deriveHierarchy(p2$reductions$PCA, ann.by.level$annotation$l3, max.depth=3) #build automatic hierarchy from the l3 of annotation_by_level
clf.tree.auto <- hierarchyToClassificationTree(hierarchy) #please do not rerun this step, 
#saveRDS(clf_tree_auto, file="~/data/separability/mouse_visp_marker_selection/clf_tree_auto.rds") 
#clf_tree_auto <- readRDS("~/data/separability/mouse_visp_marker_selection/clf_tree_auto.rds")
#plotTypeHierarchy(clf.tree.auto)
ann.by.levelM <- mergeAnnotationByLevels(ann.by.level$annotation$l3, clf.tree.auto) #contain the l3 labels from ann_by_level, while the intermediated types/subtypes are from clf_tree_auto.
```

#### Get a partly collapsed tree (the original l2 is removed, only l1 and the original l3 are left): the hierarchy will only have l0 (root), l1, and l2 (the new l2 here are the original l3 clusters)
```{r}
ann.by.levelM.collapsedl2 <- ann.by.levelM %$% list(l1,l3) %>% setNames(c("l1", "l2"))
#build a new hierarchy
hierarchyM <- ann.by.levelM %$% split(l3,l1) %>% lapply(function(n) unlist(n) %>% unique())
clf.tree.M <- hierarchyToClassificationTree(hierarchyM)
saveRDS(clf.tree.M, file="~/data/separability/mouse_visp_marker_selection/clf_treeM.rds")
#plotTypeHierarchy(clf.tree.M)
#The tree does not always look alright, some subtypes are merged together when they not are supposed to be
```

#### Manual tree (correct tree))
```{r}
clf.tree <- read_rds("~/data/separability/mouse_visp_marker_selection/clf_tree.rds")
annotation.by.level <- read_rds("~/data/separability/mouse_visp_marker_selection/annotation_by_level.rds")
```

Normalization of the normalized count data: TF-IDF (for estimating gene expression specificity [cell-wise])
```{r}
cm.norm <- p2$counts %>% Matrix::t() %>% normalizeTfIdfWithFeatures() #This step has been done, it just needs to be done once!
```

#### ML predicted tree
K Means Clustering at the l1 levels (k=2, supposed to be "Inhibitory" and "Excitatory")
```{r}
cm.norm.t <- t(cm.norm) %>% as.matrix() %>% as.data.frame() #change the tf-idf (sparse matrix) to dataframe. row: all cells, col: the 22 genes, values: tf-idf
annl3 <- ann.by.level$annotation$l3 %>% as.data.frame() # grab the annotation for all genes at l3 levels (each cell is assigned to one of 41? clusters)
colnames(annl3)<-"clusters" #names the cluster column
cm.norm.t %<>% merge(annl3, by=0) # merge by row names (by=0 or by="row.names") # merge the tf-idf and annotation data for each cell
cm.norm.t.by.clusters <- aggregate(cm.norm.t[,2:23], cm.norm.t$clusters%>%list(), mean) #calculate the mean tf-idf for each cluster and for each of the 22 genes. export: row 41? clusters, col 22 genes, value mean tf-idf among the all cells belonging to each cluster.
names(cm.norm.t.by.clusters)[names(cm.norm.t.by.clusters)=="Group.1"] <- "clusters" # rename the cluster column in the aggregated dataframe
write.csv(cm.norm.t.by.clusters, "~/data/separability/mouse_visp_marker_selection/cm_norm_clusters.csv", row.names=FALSE) # save the prepared dataframe, for K Mean clustering in python
#K Mean clustering, k=2
kmeans.label.k2 <- read.csv("~/data/separability/mouse_visp_marker_selection/kmeans_label_k2.csv") # import the predicted clustering result (K=2) 
cm.norm.t.by.clusters %<>% merge(kmeans.label.k2$l1_labels, by=0,row.names=FALSE) %<>% select(-matches("Row.names")) # merge by row names (by=0 or by="row.names"), merged the predicted labels to the dataframe, removed the "Row.names" column
names(cm.norm.t.by.clusters)[names(cm.norm.t.by.clusters)=="y"] <- "l1.label"  # rename the predicted label column
cm.norm.t.by.clusters.l1 <- cm.norm.t.by.clusters %>% split(.$"l1.label") # this list contain two dataframes, corresponding to each of the two predicted kmeans labels ("0", "1")
#Write the two dataframes within cm.norm.t.by.clusters.l1, each, to a csv file.
#prepare dataset for Affinity propogation clustering in python
for (i in cm.norm.t.by.clusters.l1 %>% names()) {
    filename = paste0("~/data/separability/mouse_visp_marker_selection/main",i,".csv")
    write.csv(cm.norm.t.by.clusters.l1[[i]], filename, row.names=FALSE)
}
```
Affinity Propagation for l2-level clustering within main neuron type "0"
```{r}
cm.norm.t.by.clusters.l1l2 <- cm.norm.t.by.clusters.l1 
#Affinity propogation clustering in python
APlabel.main0.l2 <- read.csv("~/data/separability/mouse_visp_marker_selection/APlabel_main0_l2.csv") # import the clustering result
APlabel.main0.l2 %<>% sapply(function(n) paste0("main0_",n)) %>% as.data.frame() # rename the cluster names
cm.norm.t.by.clusters.l1l2$`0` %<>% cbind("l2.label"=APlabel.main0.l2$main0_l2_labels) %<>% select(-matches("Row.names")) # merge the main0-clusters into the dataframe cm.norm.t.by.clusters.l1l2 
#caution, if there are NA, it may make the order the clusters differ, one should be very careful when combinding two dataset, maybe try cbind(x,y,by=row.names)
names(cm.norm.t.by.clusters.l1l2$`0`)[names(cm.norm.t.by.clusters.l1l2$`0`)=="y"] <- "APlabel_main0_l2" #name the l2 cluster column
```
Affinity Propagation for l2-level clustering within main neuron type "1" 
```{r}
APlabel.main1.l2 <- read.csv("~/data/separability/mouse_visp_marker_selection/APlabel_main1_l2.csv") 
APlabel.main1.l2 %<>% sapply(function(n) paste0("main1_",n)) %>% as.data.frame()
cm.norm.t.by.clusters.l1l2$`1` %<>% cbind("l2.label"=APlabel.main1.l2$main1_l2_labels) %<>% select(-matches("Row.names"))
names(cm.norm.t.by.clusters.l1l2$`1`)[names(cm.norm.t.by.clusters.l1l2$`1`)=="y"] <- "APlabel_main1_l2"
```
Construct the predicted tree
```{r}
hierarchy.pre <- lapply(cm.norm.t.by.clusters.l1l2, function(n) split(n$clusters,n$l2.label)%>%lapply(as.character)) #manually build up the tree
clf.tree.pre <- hierarchyToClassificationTree(hierarchy.pre) 
saveRDS(clf.tree.pre, "~/data/separability/mouse_visp_marker_selection/clf_treePre.csv")
#png("/Users/yuanli/Documents/Degree_project/Hierarchical_annotation/bin/data/separability/mouse_visp_marker_selection/clf_treePre.png", , width = 480, height = 1200)
#plotTypeHierarchy(clf.tree.pre)
#dev.off()
#clf_treePre <- readRDS("~/data/separability/mouse_visp_marker_selection/clf_treePre.csv")
ann.by.levelM.pre <- mergeAnnotationByLevels(ann.by.level$annotation$l3, clf.tree.pre)
```

## Get the annotation information at the higher levels of the newly built hierarchy
```{r, message=FALSE, results="hide"}
ann.all.type.auto <- getAnnotationAllType(ann.by.levelM) #for automatic tree
ann.all.type.coll2 <- getAnnotationAllType(ann.by.levelM.collapsedl2) #for the l2-removed automatic tree
ann.all.type.man <- getAnnotationAllType(annotation.by.level) #for the correct manual tree
ann.all.type.pre <- getAnnotationAllType(ann.by.levelM.pre) #for the ML predicted tree
```

## Get DE genes and pre-select markers from them
```{r, message=FALSE, results="hide"}
pre.selected.markers.auto <- getPreMarkers(p2, ann.all.type.auto)
pre.selected.markers.coll2 <- getPreMarkers(p2, ann.all.type.coll2)
pre.selected.markers.man <- getPreMarkers(p2, ann.all.type.man)
pre.selected.markers.pre <- getPreMarkers(p2, ann.all.type.pre)
```

## Select markers from pre-markers
```{r, message=FALSE, results="hide"}
marker.list.auto <- selectMarkerAllTypes(pre.selected.markers.auto, cm.norm, ann.all.type.auto,
                                     max.iters=500, max.uncertainty=0.25, log.step=5, verbose=1, n.cores=10)
saveRDS(marker.list.auto, file="~/CellAnnotatoR/notebooks/markersfiles/marker.list.auto.rds")
marker.list.auto %>% .[sapply(., function(x) length(x$expressed) > 0)] %>% 
  markerListToMarkup(file="~/CellAnnotatoR/notebooks/markersfiles/visp_markers_auto.txt", group.by.parent=F)

marker.list.coll2 <- selectMarkerAllTypes(pre.selected.markers.coll2, cm.norm, ann.all.type.coll2,
                                     max.iters=500, max.uncertainty=0.25, log.step=5, verbose=1, n.cores=10)
saveRDS(marker.list.coll2, file="~/CellAnnotatoR/notebooks/markersfiles/marker.list.coll2.rds")
marker.list.coll2 %>% .[sapply(., function(x) length(x$expressed) > 0)] %>% 
  markerListToMarkup(file="~/CellAnnotatoR/notebooks/markersfiles/visp_markers_coll2.txt", group.by.parent=F)

marker.list.man <- selectMarkerAllTypes(pre.selected.markers.man, cm.norm, ann.all.type.man,
                                     max.iters=500, max.uncertainty=0.25, log.step=5, verbose=1, n.cores=10)
saveRDS(marker.list.man, file="~/CellAnnotatoR/notebooks/markersfiles/marker.list.man.rds")
marker.list.man %>% .[sapply(., function(x) length(x$expressed) > 0)] %>% 
  markerListToMarkup(file="~/CellAnnotatoR/notebooks/markersfiles/visp_markers_man.txt", group.by.parent=F)

marker.list.pre <- selectMarkerAllTypes(pre.selected.markers.pre, cm.norm, ann.all.type.pre,
                                     max.iters=500, max.uncertainty=0.25, log.step=5, verbose=1, n.cores=10)
saveRDS(marker.list.pre, file="~/CellAnnotatoR/notebooks/markersfiles/marker.list.pre")
marker.list.pre %>% .[sapply(., function(x) length(x$expressed) > 0)] %>% 
  markerListToMarkup(file="~/CellAnnotatoR/notebooks/markersfiles/visp_markers_pre.txt", group.by.parent=F)

```

## Re-annotation
```{r}
MARKER <- parseMarkerFile("~/CellAnnotatoR/notebooks/markersfiles/visp_markers_auto.txt")
parents <- sapply(MARKER, function(x) x$parent) %>% unique() %>% setdiff(c("root",names(MARKER))) #get parents that was removed in the previous steps due to their lack of markers. 
MARKER %<>% .[sapply(., function(x) !(x$parent %in% parents))] #remove cell types that have parent in parents.
clf.data.auto <- getClassificationData(cm.norm, MARKER, prenormalized=T)
saveRDS(clf.data.auto, "~/CellAnnotatoR/notebooks/classificationtrees/clf.data.auto.rds")
ann.by.level.auto <- assignCellsByScores(NULL, clf.data.auto, clusters=p2$clusters$PCA$leiden) # re-annotation without graph
#ann.by.level.auto.g <- assignCellsByScores(p2$graphs$PCA, clf.data.auto, clusters=p2$clusters$PCA$leiden) # re-annotation with graph
#saveRDS(ann.by.level.auto.g, file="~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.auto.g.rds")
ann.by.level.auto.g <- readRDS("~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.auto.g.rds")

MARKER <- parseMarkerFile("~/CellAnnotatoR/notebooks/markersfiles/visp_markers_coll2.txt")
parents <- sapply(MARKER, function(x) x$parent) %>% unique() %>% setdiff(c("root",names(MARKER)))
MARKER %<>% .[sapply(., function(x) !(x$parent %in% parents))]
clf.data.coll2 <- getClassificationData(cm.norm, MARKER, prenormalized=T)
saveRDS(clf.data.coll2, "~/CellAnnotatoR/notebooks/classificationtrees/clf.data.coll2.rds")
ann.by.level.coll2 <- assignCellsByScores(NULL, clf.data.coll2, clusters=p2$clusters$PCA$leiden)
#ann.by.level.coll2.g <- assignCellsByScores(p2$graphs$PCA, clf.data.coll2, clusters=p2$clusters$PCA$leiden)
#saveRDS(ann.by.level.coll2.g, file="~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.coll2.g.rds")
ann.by.level.coll2.g <- readRDS("~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.coll2.g.rds")

clf.data.man <- getClassificationData(cm.norm, "~/CellAnnotatoR/notebooks/markersfiles/visp_markers_man.txt", prenormalized=T)
saveRDS(clf.data.man, "~/CellAnnotatoR/notebooks/classificationtrees/clf.data.man.rds")
ann.by.level.man <- assignCellsByScores(NULL, clf.data.man, clusters=p2$clusters$PCA$leiden)
#ann.by.level.man.g <- assignCellsByScores(p2$graphs$PCA, clf.data.man, clusters=p2$clusters$PCA$leiden)
#saveRDS(ann.by.level.man.g, file="~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.man.g.rds")
ann.by.level.man <- readRDS("~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.man.g.rds")

clf.data.pre <- getClassificationData(cm.norm, "~/CellAnnotatoR/notebooks/markersfiles/visp_markers_pre.txt", prenormalized=T)
saveRDS(clf.data.pre, "~/CellAnnotatoR/notebooks/classificationtrees/clf.data.pre.rds")
ann.by.level.pre <- assignCellsByScores(NULL, clf.data.pre, clusters=p2$clusters$PCA$leiden)
#ann.by.level.pre.g <- assignCellsByScores(p2$graphs$PCA, clf.data.pre, clusters=p2$clusters$PCA$leiden)
#saveRDS(ann.by.level.pre.g, file="~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.pre.g.rds")
ann.by.level.pre.g <- readRDS("~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.pre.g.rds")
```

## QC: positive, negative and coverage: mean uncertainty per level

### For partly collapsed tree
#### ***** Without graph  *****
```{r}
score.info.coll2 <- getMarkerScoreInfo(clf.data.coll2)
unc.per.cell.coll2 <- scoreCellUncertaintyPerLevel(ann.by.level.coll2, score.info.coll2)
unc.per.cell.coll2 %>% lapply(function(n) sapply(n, mean))
```
#### ***** With graph  *****
```{r}
score.info.coll2.g <- getMarkerScoreInfo(clf.data.coll2)
unc.per.cell.coll2.g <- scoreCellUncertaintyPerLevel(ann.by.level.coll2.g, score.info.coll2.g)
saveRDS(unc.per.cell.coll2.g, "~/CellAnnotatoR/notebooks/uncertainty/unc.per.cell.coll2.g.rds")
unc.per.cell.coll2.g %>% lapply(function(n) sapply(n, mean))
```

### For the correct manual tree:
#### ***** Without graph  *****
```{r}
score.info.man <- getMarkerScoreInfo(clf.data.man)
unc.per.cell.man <- scoreCellUncertaintyPerLevel(ann.by.level.man, score.info.man)
unc.per.cell.man %>% lapply(function(n) sapply(n, mean))
```
##### With Viktor's files
#### ***** Without graph  *****
```{r}
clf.data <- getClassificationData(cm.norm, "~/data/separability/mouse_visp_marker_selection/visp_markers_auto.txt", prenormalized=T)
ann.by.level <- assignCellsByScores(NULL, clf.data, clusters=p2$clusters$PCA$leiden)
score.info <- getMarkerScoreInfo(clf.data)
unc.per.cell.man <- scoreCellUncertaintyPerLevel(ann.by.level, score.info)
unc.per.cell.man %>% lapply(function(n) sapply(n, mean))
```
##### With Viktor's files
#### ***** With graph  *****
```{r}
ann.by.level.e <- assignCellsByScores(p2$graphs$PCA, clf.data, clusters=p2$clusters$PCA$leiden)
#saveRDS(ann.by.level.e, file="~/CellAnnotatoR/notebooks/annbylevels/ann.by.level.man.g.rds")
score.info.e <- getMarkerScoreInfo(clf.data)
unc.per.cell.man.e <- scoreCellUncertaintyPerLevel(ann.by.level.e, score.info.e)
saveRDS(unc.per.cell.man.e, "~/CellAnnotatoR/notebooks/uncertainty/unc.per.cell.man.g.rds")
unc.per.cell.man.e %>% lapply(function(n) sapply(n, mean))
```
### For the automatic tree:
#### ***** Without graph  *****
```{r}
score.info.auto <- getMarkerScoreInfo(clf.data.auto)
unc.per.cell.auto <- scoreCellUncertaintyPerLevel(ann.by.level.auto, score.info.auto)
unc.per.cell.auto %>% lapply(function(n) sapply(n, mean))
```
#### ***** With graph  *****
```{r}
score.info.auto.g <- getMarkerScoreInfo(clf.data.auto)
unc.per.cell.auto.g <- scoreCellUncertaintyPerLevel(ann.by.level.auto.g, score.info.auto.g)
saveRDS(unc.per.cell.auto.g, "~/CellAnnotatoR/notebooks/uncertainty/unc.per.cell.auto.g.rds")
unc.per.cell.auto.g %>% lapply(function(n) sapply(n, mean))
```

### For the predicted tree:
#### ***** Without graph  *****
```{r}
score.info.pre <- getMarkerScoreInfo(clf.data.pre)
unc.per.cell.pre <- scoreCellUncertaintyPerLevel(ann.by.level.pre, score.info.pre)
unc.per.cell.pre %>% lapply(function(n) sapply(n, mean))
```
#### ***** With graph  *****
```{r}
score.info.pre.g <- getMarkerScoreInfo(clf.data.pre)
unc.per.cell.pre.g <- scoreCellUncertaintyPerLevel(ann.by.level.pre.g, score.info.pre.g)
saveRDS(unc.per.cell.pre.g, "~/CellAnnotatoR/notebooks/uncertainty/unc.per.cell.pre.g.rds")
unc.per.cell.pre.g %>% lapply(function(n) sapply(n, mean))
```

