---
title: "Analysis of marmoset data"
output: html_document
---

This file shows analysis of the visual cortex (V1) data from marmoset from the [CellAnnotatoR tutorial](https://github.com/khodosevichlab/CellAnnotatoR/blob/master/vignettes/mca_marker_selection.md) 

```{r}
library(ggplot2)
library(magrittr)
library(Matrix)
library(pbapply)
library(dplyr)
library(pagoda2)
library(CellAnnotatoR)

theme_set(theme_bw())
```

The visual cortex (V1) data from marmoset has been re-processed

## Import the visual cortex (V1) data from marmoset

```{r}
p3 <- readRDS("~/data/marmoset/marm022_v1_p2.rds")
names(p3)
```

## Run clustering and visualize.

```{r}
p3$getKnnClusters(type="PCA", method=conos::leiden.community, n.iterations=10, 
                  resolution=6, name="leiden")
conos::embeddingPlot(p3$embeddings$PCA$UMAP_graph, groups=p3$clusters$PCA$leiden)
```

## Define hierarchy

The automated algorithm is rather simple and will be improved in future. And currently, it can not be used on the aligned data.

```{r} 
hierarchy3 <- deriveHierarchy(p3$reductions$PCA, p3$clusters$PCA$leiden, max.depth=3)
clf_tree3 <- hierarchyToClassificationTree(hierarchy3)
plotTypeHierarchy(clf_tree3)
```

Annotation by level:

```{r}
ann_by_level3 <- mergeAnnotationByLevels(p3$clusters$PCA$leiden, clf_tree3)
plotAnnotationByLevels(p3$embeddings$PCA, ann_by_level3, font.size=c(2, 3), n.col=3, shuffle.colors=T)
```

To run render the file: rmarkdown::render(marmoset.RMD)